Hierarchical Time Series Forecasting in Emergency Medical Services

### 2023-05-10 journal_correspondence/response_letter/response_letter_JSR.qmd

---
title: "Response letter"
format: pdf
---

Dear Sir or Madam,

Please find enclosed a revised version of our paper submitted for peer review for the Journal of Service Research. Your comments have been very much appreciated and we have tried to suitably modify/amend our paper for those comments to be reflected in the revised version.

We quote below the detailed referees’ comments that are followed by our response. In all cases, we point out, where necessary, the corresponding amendments in the new version of the paper and highlighted them in blue.

# Editor comment

Manuscript ID JSR-22-468 entitled "Hierarchical Time Series Forecasting in Emergency Medical Services" which you submitted to the Journal of Service Research, has been reviewed. I read the manuscript and assigned it to an Associate Editor (AE). The AE selected the reviewers. Three expert reviewers responded with detailed and constructive comments. The AE wrote a summary letter with additional comments. I read all the reports and reread the manuscript. The reviewer and AE reports are located below.

With the exception of R3, the review team finds the research topic to be very important and sees the potential for your manuscript to have substantial impact. I also like the topic and see similar potential. The AE and I think the issues raised by R3 can be addressed in a revision.

R1, R2 and the AE raised many clarifying questions and had concrete suggestions for improve the work. I ask that you give serious consideration to these comments. The current MS is 33 pages, and we allow up to about 50 pages, and so you should have enough space to add the clarifications, possibly as an appendix.

I agree with many of R3's comments. JSR is not a methods journal, and the readers will generally be interested in substantive implications of the work. The evaluation currently focuses on several accuracy metrics such as MACE and MSSE, but what does improved forecasting accuracy mean to business metrics? Perhaps you could do some calculations to demonstrate the importance of your research in terms of business outcomes? Another way to address this is to give more detail about the natural of the decisions that are being made. How specifically can a planner “reconfigure units” and what sorts of options are available in “redeploying staff”? Help the reader understand the managerial decisions being made.

R3 also raises an issue about the generalizability. The AE has some suggestions for broadening the introduction slightly, and there could be a bit more discussion about other possible contexts in which your model might be used. R3 gives two cites that seem to address the problem; I am interested in whether these methods already solve your problem, or whether they would require substantial modification.

I liked your point about reproducibility very much. It is great that you will released you code on Github. Would it be possible to release your data as a benchmark data set for the research community? I understand that there may be NDAs involved, but publicly available data sets help the field advance.

This is minor, but there seem to be issues with your bibliography software. Sometimes it inserts initials in the text citations, other times not. Sometimes commas are missing, e.g., "Haugsbø Hermansen and Mengshoel (2021)". There are other writing errors, e.g., sometimes "neural network" is capitalized and other times not. It is not a proper noun.

# Associate Editor

_General comments to author: This manuscript looks at how to best forecast ambulance demand. These forecasts are crucial for planning and deploying staff and fleet. They are required at different levels (e.g., national, regional, and sub-regional levels) and are often made independently of each. So, the resulting forecasts at the different levels could be inconsistent. To avoid this problem, the authors exploit the hierarchical and grouped structure of the demand time series and apply several forecasting methods. In their empirical study with daily incident data from an ambulance service in Great Britain from October 2015 to July 2019, an ensemble of forecasting models yields the best forecasts that are also coherent at all levels._

_We received three very excellent reviews that came up with mixed recommendations. Still, they share your excitement about the topic. They are mostly consistent in recognizing the manuscript’s strengths and weaknesses. All three reviewers read the paper as a “forecasting exercise”. Instead, you need to better highlight the substantive insights that your “forecasting exercise” provides. In addition to the recommendations of the reviewers, my recommendation below might help you achieve this aim._

_All reviewers have questions and recommendations regarding the details of your time series. They seem to be justified and not in conflict with each other. So, consider implementing them._

_All reviewers have questions regarding your look “beyond the point estimate”. Consider them. It might help to also come up with an example that outlines the advantage of going beyond point estimates in greater detail._

_R1 has a long list of precise and helpful suggestions for improving the manuscript. It looks like they are all justified, and it should be manageable to implement them. The same holds for R2 and R3. R3’s remark regarding Figure 4 and the performance of the naïve model that seems to be not much worse is particularly crucial._

Response: Thank you for your feedback and for summarizing the main points from the reviewers. We appreciate your suggestion to highlight the insights provided by the forecasting exercise, and we have worked on the manuscript to make this more explicit in the revised manuscript. We agree that the reviewers have raised important questions and recommendations regarding the details of our study, and we have addressed these in the revised manuscript. Regarding the recommendation to look beyond the point estimate, we have taken the reviewers' suggestions into account and provide an example that illustrates the advantage of doing so. We have implemented R1's suggestions in the revised manuscript. We have also responded to R3's remark regarding the performance of the Naïve model and Figure 4.

## AE comments

In addition to the reviewers, I recommend also addressing my following suggestions:

_Comment 1. Consider broadening the scope of the manuscript by not focusing instantly on ambulance demand. Probably, other services also share the problem of having different levels. Outline them and explain why they all face the core problem that you address. R3 makes a comparable statement (“generalizability”). Such broadening might also enable you to better link your work to previous research in the Journal of Service Research._

  **Response:** Thank you for the suggestion! We agree that broadening the scope of the study to include other services that face similar hierarchical and grouped time series forecasting issues would make it more relevant and interesting to the Journal of Service Research audience. We have added a new subsection to the manuscript in the conclusion section, where we discuss other services that share the same problem and how our approach could be applied to those services. We also linked our work to previous research in the Journal of Service Research to provide a more comprehensive understanding of our contribution.

_Comment 2 (Discuss with Rob). It would also help to have a rather detailed example, maybe even a numerical example, to better describe why separate forecasts at different levels can cause problems. Right now, it sounds plausible that such problems occur, but you are still rather vague concerning the details of these problems._

  **Response:** Thank you for your comment. We agree that providing a detailed example can better illustrate the problems that arise when separate forecasts at different levels are inconsistent. We have added a simple numerical example to the manuscript, which outlines the issue of incoherent in hierarchical forecasts.

_Comment 3: You sometimes mentioned that the data runs from 2015-2019 (e.g., in the abstract) and sometimes from 2015-2020 (page 3). Please harmonize._

  **Response:** Thank you for bringing this to our attention. We have gone through the manuscript and made sure that we consistently report the duration of the dataset as 2015 to 2019, which is the correct period

_Comment 4: Consider summarizing previous research in a table. Such a table should also enable you to better highlight your contribution (by also including your work in the table)._

  **Response:**  Thank you. We agree that a table could provide a clearer overview of the related literature and our contribution. Therefore, we have now included a table in the literature review section summarizing previous research, including our study's contribution.

_Comment 5: Make sure that all tables and figures are self-contained._

  **Response:** Thank you for your comment. We have reviewed each table and figure in the manuscript and ensure that they are complete and understandable without relying on the main text and all necessary information such as axis labels, legends, etc are provided.

_Comment 6: Table 1: Make it easier for the reader to understand all values. For example, instead of “105”, write “105 = 35 x3”, etc._

  **Response:** Thank you for your comment.We have updated Table 1 to make it easier for readers to understand all values by providing a breakdown of the calculation.

_Comment 7 (Check with Rob): If possible, display the number of observations in all tables and figures with statistical results._

  **Response:** Thank you for your comment. We agree that displaying the number of observations is important when reporting statistical results. However, in the context of forecasting, what matters more is testing the accuracy of the forecast using new data (i.e., test set) by employing time series cross-validation. Statistical significance matters more when we are interested in fitting a model to the data. Therefore, we believe that it is not necessary to display the number of observations in all tables and figures with statistical results in this manuscript.

# Reviewer 1

_**General comment:** In this study, the authors assess the efficacy of various hierarchical time series forecasting techniques within the domain of emergency medical services. Specifically, they consider the hierarchical and group structure of ambulance demand. Utilizing daily incident data from an ambulance service in Great Britain spanning from October 2015 to July 2019, the authors conduct forecasting and performance evaluations. They employ an ensemble of forecasting models, in which the predictions of multiple models are merged to produce superior results compared to any single forecasting model. The paper is commendably organized, clearly written, and easy to comprehend. However, it lacks sufficient detail on the methodology, as noted in my comments below. I encourage the authors to include more in-depth information to enhance the overall quality of the paper._

**Response:** Thank you for taking the time to review our paper. We appreciate your positive comments and understand your concerns regarding the lack of detail on the methodology. We have addressed your comments in the following responses.

_**Major Comments:** The description of the approaches in section 3 and section 4 are missing important details that would enhance the readers’ understanding of the paper._

**Response:** Thank you for pointing out that our descriptions in sections 3 and 4 could benefit from more detail. We have revised those sections to include more specific information on the forecasting models and hierarchical and grouped time series techniques utilized in our study.

_**Comment 1**: The data points used in this time series forecasting pertain to "attended incidents”, which refer to situations where a physical resource has been deployed to respond to an incident. However, this definition only accounts for actual responses, and does not consider potential failed responses. I suspect that the response rate will not always be 100% due to various reasons, such as lack of resources, and it is important to take these failed responses into account in forecasting future emergency medical services (EMS) needs. Is data on failed responses available? If not, it would be beneficial to at least acknowledge and discuss this aspect of the forecast in the analysis._

**Response:** Thank you for your thoughtful comment regarding the potential impact of failed responses on our time series forecasting. We agree that failed responses could have an impact on forecasts, as the actual demand for ambulance services could be higher than what we have observed in the attended incidents data. Unfortunately, as you noted, the dataset we used only includes information on attended incidents, and we do not have access to information on failed responses. Therefore, we were not able to directly account for the potential impact of failed responses in our analysis. However, we have acknowledged this limitation in our manuscript and discussed its potential impact on our forecasting results. We agree that it would be valuable for future research to investigate the impact of failed responses on EMS forecasting, if data on these incidents becomes available. In the meantime, we hope that our study can provide some insights into the hierarchical time series forecasting techniques that can be used with attended incident data, and how these techniques can potentially improve the accuracy of EMS demand forecasting. It is also important to note that our methodology for hierarchical time series forecasting can be applied to any time series data in EMS, including those that may include failed responses. However, the forecasting results may differ depending on the specific characteristics of the data.

_Comment 2. It is important to note the value of this research in the context of emergency medical services (EMS). Both high-level and low-level forecasts are necessary for effective decision-making in EMS. For example, control area EMS forecasts are high-level forecasts that inform strategic decisions about how to allocate limited resources to lower levels, such as health boards and stations/hospitals. These types of decisions are typically made periodically, not on a daily or weekly basis. In this context, the research presented in this paper has clear value. However, it is not clear to me what specific value the research brings at a lower level. It would be beneficial if this aspect could be further explained in the paper._

**Response:** Thank you for your insightful comment. We appreciate your point about the importance of both high-level and low-level forecasts for effective decision-making in EMS. In our study, we use a methodology that provides forecasts for all levels of a given hierarchy including high-level and low-level of the hierarchy. We believe that wherever decisions are made about the future in the hierarchy, forecasts would be used as an input. While you pointed out values on high-level forecasts, but forecasts are also valuable in at lower levels for shorter-term decision-making and planning. For example, hospitals or ambulance stations could use such forecasts to plan for staffing and resource allocation, ambulance dispatching, staff-to-shift assignment, staff rescheduling based on the anticipated volume and type of incidents. Additionally, generating forecasts at lower levels could potentially improve the accuracy of the high-level forecasts, by providing more detailed information on the nature and priority of incidents. This could help to identify patterns and trends in demand that may not be apparent at the higher level.

_Comment 3. Second paragraph of page 9 mentions “the strength of trend” and “the strength of seasonality”. Can you explain what those are or how they are calculated?_

**Response:** Thank you for the comment. In this paper, the strength of trend and seasonality were calculated using the "STL" (Seasonal and Trend decomposition using Loess) decomposition method, which is a widely used and flexible method for decomposing time series data into trend, seasonal, and remainder components. The decomposition of a time series  $y_t$ is written as $y_t = T_t + S_{t} + R_t$, where $T_t$ is the smoothed trend component, $S_t$ is the seasonal component and $R_t$ is a remainder component. The strength of trend is defined as:

$$F_T = \max\left(0, 1 - \frac{\text{Var}(R_t)}{\text{Var}(T_t+R_t)}\right)$$

For strongly trended data, the seasonally adjusted data should have much more variation than the remainder component. Therefore Var($R_t$)/Var($T_t+R_t$) should be relatively small. But for data with little or no trend, the two variances should be approximately the same.

The strength of seasonality is defined similarly:

$$F_S = \max\left(0, 1 - \frac{\text{Var}(R_t)}{\text{Var}(S_{t}+R_t)}\right).$$

 series with seasonal strength  $F_S$, close to 0 exhibits almost no seasonality, while a series with strong seasonality will have $F_S$ close to 1 because Var($R_t$) will be much smaller than Var($S_t+R_t$).

_Comment 4. One of my main concerns is with regard to section 3.2. The current discussions of the forecasting approaches are inadequate in terms of important details. To improve the understanding of the paper, please provide a more detailed description of each of these approaches._

**Response:** Thank you for your feedback. We appreciate your concern regarding the level of detail provided in Section 3.2 on the forecasting approaches used in this study. While we agree that it is important to provide adequate information in the paper itself, we also understand that some readers may be interested in more technical details and code implementation. For this reason, we have provided a Github repository with all technical details and R codes for the forecasting approaches used in this study. However, we acknowledge that more information can be provided in the paper itself to enhance readers' understanding. To address your concern, we have revised Section 3.2 to provide a more detailed description of each of the forecasting approaches used, including the mathematical framework, assumptions, and key parameters. We hope that this revised section, along with the Github repository, will provide readers with the necessary information to understand and replicate the methods used in this study.

_Comment 4.1: In this research, the ”naive forecasting approach” is used as a benchmark. Is this the forecasting approach currently used in practice? If not, why was it selected as the benchmark? I expect that more rigorous forecasting approaches will outperform the naive approach. It would be more compelling to demonstrate the real-world value of your approach if it were compared against the approach currently used in practice._

**Response:** Thank you for the suggestion. We appreciate your feedback. We would like to clarify that our intention in this paper is not to provide an exhaustive list of forecasting methods. Instead, we aim to provide a methodology that can incorporate forecasts produced by any method. Our focus is on developing a hierarchical time series forecasting approach that takes into account the hierarchical and group structure of ambulance demand. We have selected the naive method as a benchmark because it is widely used in forecasting literature and practice and is a simple and easily understandable method for users.

Forecasts are often used as input for other decision-making systems that often involve simulation, and it is quite common to use the empirical distribution of the demand as a forecast in simulations. Also, the naive method can be surprisingly accurate. Therefore, if any forecasting method should be able to provide more accurate results than naive, it would justify its use in practice, if not there is no need to use more complicated methods.

_Comment 4.2: In your discussion of ETS, you mention that the trend, seasonality, and error terms can take various forms, such as additive, multiplicative, or mixed. Can you specify which form you used for each of these components in your research and explain the reasoning behind your choice?_

**Response:** Thank you for the question. In our research, we used an automatic algorithm to determine the appropriate form for the trend, seasonality, and error terms in each time series. Specifically, we used the ETS() function in the fable package of R, which selects the most appropriate model for each time series based on statistical criteria such as Akaike information criterion (AIC). Since we had a large number of time series (1530), it was not practical to manually select the appropriate form for each component for each time series. Instead, the automatic algorithm selects the most appropriate components based on the characteristics of each individual time series. As a result, a mix of different component forms (additive or multiplicative) were used, depending on the specific characteristics of each time series.

_Comment 4.3 (Rob): In your research, the Generalized Linear Model (GLM) approach is used for forecasting. Can you provide the equation used for this approach in a concise form? Additionally, in your description of the Fourier terms, it is mentioned that they are used to capture yearly seasonality. Does your research also take into account monthly seasonality and if so, how is it captured? Furthermore, for yearly seasonality, what is the specific form of the Fourier term used?_

**Response:**

_Comment 4.4: Can you explain what “serial dependence” means in the TSGLM model in the context of EMS forecasting?_

**Response:** Thank you for the comment. In the context of EMS forecasting, serial dependence refers to the situations where ambulance demand at one point in time are correlated with the demand at previous points in time. In other words, the past values of the time series influence the current values, and this influence can be modeled using a variety of techniques. The TSGLM model, for example, explicitly accounts for serial dependence by including lagged values of the ambulance demand in the model. This is important in EMS forecasting because it allows the model to capture patterns and trends in the data that are dependent on the history of the time series.

_Comment 4.5 (Rob): It is clear that the naive method performed poorly in the results. Can you explain the reasoning behind its inclusion in the ensemble method? Have you also evaluated the performance of an ensemble method that combines forecasts from only ETS, GLM and TSGLM, without the naive method? I would be interested in seeing the results from this alternative ensemble method._

**Response:**

_Comment 5 (Rob to check the response): In the first paragraph of section 3.3, you stated that the model development and hyper-parameter tuning process is conducted using the training data alone. Is this a commonly employed approach within the literature? From my understanding, when hyper-parameter tuning is involved, a dataset is typically divided into three sets: training, validation, and test sets. The training set is used to develop the model, the validation set is employed for tuning the parameters, and the test set is utilized to report the error measurement metric. Can you provide further clarification on this methodology and how it compares to the standard approach in the literature?_

**Response:** Thank you for your comment and question. We agree that traditionally, datasets are divided into three sets: training, validation, and test sets, and the validation set is used to tune the hyperparameters of the model. However, please note that we employ time series cross-validation to assess the forecast accuracy, and thus the use of a separate "validation dataset" becomes unnecessary when estimating the model parameters through cross-validation with the training data.Time series cross-validation involves dividing the data into multiple training and testing sets based on the time series structure. Each testing set consists of a number of observations equal to the forecast horizon, and the corresponding training set only includes observations that occurred before the test set. This ensures that future observations are not used in constructing the forecast, making it more robust to evaluate forecast accuracy on new data. Therefore, we believe that our methodology is appropriate for time series forecasting and provides a more robust approach for evaluating forecast accuracy on new data.

_Comment 6: In Section 3.3, it is not clear what the forecasting horizon j represents. The unit of time, whether it is a day, a week, or multiple weeks, is not specified and this makes it difficult to understand the logic behind the point forecast error e_j . Additionally, it would be helpful to explain why the scale-independent measures such as MASE and MSSE were chosen and how it relates to other commonly used accuracy measures._

**Response:** Thank you for your comment. To clarify, the forecasting horizon j represents the number of time periods ahead that we are forecasting, which is j= 1,2, ..., 84 days in our study. We will make sure to include this information in the revised version of the paper.

### 2023-05-10 journal_correspondence/cover_letter.qmd
response letter - reposne to AE and reviewer 1

---
author:
  - name: Bahman Rostami Tabar & Rob&nbsp;J.&nbsp;Hyndman
    affiliation:
      - name: Bahman Rostami-Tabar
        department: Cardiff Business School
        address: 3, Colum Drive
        city: Cardiff
        country: United Kingdom
        postal-code: CF10 3EU
address:
  - Prof. Donald M. Yealy
  - Editor-in-Chief
  - Annals of Emergency Medicine
subject: Paper submission for peer review
opening: "Dear Prof. Yealy,"
closing: "Yours sincerely"
format: letter-pdf
---

We would be grateful if you could consider for publication our manuscript entitled _Hierarchical Time Series Forecasting in Emergency Medical Services_ as an Original Research article in the Annals of Emergency Medicine.

>>>>

---
author:
  - name: Bahman Rostami Tabar & Rob&nbsp;J.&nbsp;Hyndman
    affiliation:
      - name: Bahman Rostami-Tabar
        department: Cardiff Business School
        address: 3, Colum Drive
        city: Cardiff
        country: United Kingdom
        postal-code: CF10 3EU
address:
  - Prof. Ming-Hui Huang
  - Editor-in-Chief
  -  Journal of Service Research
subject: Paper submission for peer review
opening: "Dear Prof. Huang,"
closing: "Yours sincerely"
format: letter-pdf
---

We would be grateful if you could consider for publication our manuscript entitled _Hierarchical Time Series Forecasting in Emergency Medical Services_ as an Original Research article in the Journal of Service Research.

### 2023-05-10 manuscript/main.qmd
Some final changes
MS word document generated for submission, plus typos corrected
update the Github link
response letter - reposne to AE and reviewer 1

---
title: "Hierarchical Time Series Forecasting in Emergency Medical Services"
author:
  - name: Bahman Rostami-Tabar
    email: rostami-tabarb@cardiff.ac.uk
    affiliations:
        - id: carbs
          name: Cardiff University
          department: Cardiff Business School
          country: United Kingdom
          postal-code: CF10 3EU
    attributes:
        corresponding: true
  - name: Rob J. Hyndman
    email: Rob.Hyndman@monash.edu
    affiliations:
        - id: monash
          name: Monash University
          department: Department of Econometrics and Business Statistics
          country: Australia
          postal-code: VIC 3800
abstract: |
  Accurate forecasts of ambulance demand are crucial inputs when planning and deploying staff and fleet. Such demand forecasts are required at national, regional, and sub-regional levels, and must take account of the nature of incidents and their priorities. These forecasts are often generated independently by different teams within the organization. As a result, forecasts at different levels may be inconsistent, resulting in conflicting decisions and a lack of coherent coordination in the service. To address this issue, we exploit the hierarchical and grouped structure of the demand time series and apply forecast reconciliation methods to generate both point and probabilistic forecasts that are coherent and use all the available data at all levels of disaggregation. The methods are applied to daily incident data from an ambulance service in Great Britain, from October 2015 to July 2019, disaggregated by nature of incident, priority, managing health board, and control area. We use an ensemble of forecasting models and show that the resulting forecasts are better than any individual forecasting model. We validate the forecasting approach using time-series cross-validation.
keywords:
  - healthcare
  - emergency services
  - forecast reconciliation
  - ambulance demand
  - regression
bibliography: bibliography.bib
execute:
  echo: false
  message: false
  warning: false
  cache: true
pdf-engine: pdflatex
prefer-html: true
format:
  html:
    number-sections: true
  docx:
    number-sections: true
  elsevier-pdf:
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{todonotes,mathtools,bm,amsmath,mathpazo}
        \mathtoolsset{showonlyrefs}
        \setlength{\parindent}{0cm}
    journal:
      name: Annals of Emergency Medicine
      formatting: preprint
      model: 3p
      cite-style: authoryear
---

@@

>>>>

---
title: "Hierarchical Time Series Forecasting in Emergency Medical Services"
author:
  - name: Bahman Rostami-Tabar
    email: rostami-tabarb@cardiff.ac.uk
    affiliations:
        - id: carbs
          name: Cardiff University
          department: Cardiff Business School
          country: United Kingdom
          postal-code: CF10 3EU
    attributes:
        corresponding: true
  - name: Rob J. Hyndman
    email: Rob.Hyndman@monash.edu
    affiliations:
        - id: monash
          name: Monash University
          department: Department of Econometrics and Business Statistics
          country: Australia
          postal-code: VIC 3800
abstract: |
  Accurate forecasts of ambulance demand are crucial inputs when planning and deploying staff and fleet. Such demand forecasts are required at national, regional, and sub-regional levels, and must take account of the nature of incidents and their priorities. These forecasts are often generated independently by different teams within the organization. As a result, forecasts at different levels may be inconsistent, resulting in conflicting decisions and a lack of coherent coordination in the service. To address this issue, we exploit the hierarchical and grouped structure of the demand time series and apply forecast reconciliation methods to generate both point and probabilistic forecasts that are coherent and use all the available data at all levels of disaggregation. The methods are applied to daily incident data from an ambulance service in Great Britain, from October 2015 to July 2019, disaggregated by nature of incident, priority, managing health board, and control area. We use an ensemble of forecasting models and show that the resulting forecasts are better than any individual forecasting model. We validate the forecasting approach using time series cross validation.
keywords:
  - healthcare
  - emergency services
  - forecast reconciliation
  - ambulance demand
  - regression
bibliography: bibliography.bib
execute:
  echo: false
  message: false
  warning: false
  cache: true
pdf-engine: pdflatex
prefer-html: true
format:
  html:
    number-sections: true
  docx:
    number-sections: true
  elsevier-pdf:
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{todonotes,mathtools,bm,amsmath,mathpazo}
        \mathtoolsset{showonlyrefs}
        \setlength{\parindent}{0cm}
    journal:
      name: Journal of Service Research
      formatting: preprint
      model: 3p
      cite-style: authoryear
---

@@

# Reproducibility {.unnumbered}

R code to produce all results in this paper is available at
